---
title: "Bayesian dynamics recipes (ild_brms)"
author: "tidyILD authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian dynamics recipes (ild_brms)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This article gives **copy-paste templates** for modest **dynamics** extensions with [`ild_brms()`]. It is **not** a substitute for identification work, prior sensitivity, or specialized **DSEM** software when that is your estimand. See `vignette("temporal-dynamics-model-choice", package = "tidyILD")` for how these pieces fit the bigger picture.

## Prerequisites

- Packages: **brms**, **Stan** backend (**rstan** or **cmdstanr**).
- Data prepared with [`ild_prepare()`]; lag columns from [`ild_lag()`] or [`ild_panel_lag_prepare()`].

## Recipe 1: Random slope for a lagged predictor

```{r recipe-random-lag-slope}
library(tidyILD)
set.seed(1)
d <- ild_simulate(n_id = 15, n_obs_per = 12, seed = 1)
d$x <- rnorm(nrow(d))
x <- ild_prepare(d, id = "id", time = "time")
x <- ild_lag(x, x, n = 1L, mode = "gap_aware")
tpl <- ild_brms_dynamics_formula("y", "x_lag1", id_var = "id")
tpl$formula
tpl$notes
# Short chains for illustration only:
# fit <- ild_brms(tpl$formula, data = x, iter = 500, chains = 2, refresh = 0)
# ild_tidy(fit)
```

`ild_brms_dynamics_formula()` returns a **suggested** formula `(lag | id) + (1|id)`; adjust priors via [`ild_prior_ild()`] and check [`ild_diagnose()`] (R-hat, ESS).

## Recipe 2: Bivariate lag sketch (two variables)

Build two lag columns, then include both fixed effects (and optionally random slopes). A full **panel VAR** requires more structure than this template; this is a **minimal** associational extension.

```{r recipe-two-lags}
x2 <- ild_panel_lag_prepare(x, c("x", "y"), n = c(1L, 1L), mode = "gap_aware")
names(x2$data)
# Example fixed structure only (not run):
# f2 <- y ~ x_lag1 + y_lag1 + (1 | id)
# fit2 <- ild_brms(f2, data = x2$data, iter = 500, chains = 2, refresh = 0)
```

## See also

- `?ild_brms`, `?ild_prior_ild`, `?ild_diagnose`
- `vignette("temporal-dynamics-model-choice", package = "tidyILD")`

```{r session-info, echo = FALSE}
sessionInfo()
```
